Parsing Postfix log files

### 2007-08-15 doc/logparser.tex
Second pass through the corrections; all the small stuff is done. Promoted database schema subsection to a full section. Merged LMA review into prior art section, and reworked the combination somewhat.

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
\DeclareUrlCommand\urlwrap{}
\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib,nottoc]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{%
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}

@@

Section~\ref{rules} discusses the parsing rules in detail, explaining their
structure, using an example rule and sample data it matches successfully
against, followed by a discussion of rule efficiency concerns.

@@

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are already
        lots of readily available resources from which to learn.  Although
        every vendor implements a different dialect of SQL, the basics are
        the same everywhere.

@@

\subsection{Database schema}
\label{database schema}

XXX SHOULD THIS SECTION BE MOVED ELSEWHERE\@?

\subsubsection{Connections table}

\subsubsection{Results table}

@@

XXX REWRITE THIS

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.

@@

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from this parser: it attempts to present correlated log
data in a form suitable for a systems administrator to search using
grep(1), producing a CSV file\footnote{Comma-Separated Value, a very simple
format where a record is stored in a single line, with fields separated by
a comma or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.} which can subsequently be imported into a
MySQL or Berkeley DB database.  The data stored is limited to time and
date, hostname and IP address of client, sender and recipient addresses,
and the SMTP code.  The decision to support both CSV and Berkeley DB
appears to have been a limiting factor: the values which can be stored are
limited, it's difficult to link different records (foreign keys in SQL
enable this easily), and ad-hoc queries require extensive code to be
written.  Berkeley DB is not an SQL database, it only supports (key, value)
pairs; in the main table the value used is a CSV line, and the key is an
integer referred to by other tables.  The secondary by-sender,
by-recipient, by-date and by-IP tables use the sender/recipient/date/IP as
the key, and a CSV list of integers referring to the main table is used as
the value.  In effect this re-implements SQL foreign keys, but without the
full functionality offered by even the most basic of SQL databases.  It
also requires custom code to search on some combination of the above.

Handling of multiple recipients, SMTP codes or remote servers\footnote{A
single mail may be sent to multiple remote servers if it was addressed to
recipients in different domains, or Postfix needs to try multiple servers
for one or more recipients.} is not explained in the paper, but the options
appear to be:

@@

@@

While the rules have more lines of code than the algorithm, the rules are
quite simple and each rule is completely independent of its fellows.  The
algorithm is significantly more complicated and highly internally
interdependent.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\ref{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if there haven't been any
rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

    \item This is followed by rejection of sender, recipients, client
        address, etc. (action: rejection; flowchart:~4); acceptance of one
        or more mails (action: clone; flowchart:~5); or some interleaving
        of both.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections.  See
section~\ref{tracking re-injected mail} for a full discussion.

Internally generated mails lack any explicit origin (in Postfix 2.2.x) and
can only be detected using heuristics (see
section~\ref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing, local or remote delivery.  All
four are handled in exactly the same way:

@@

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from memory (flowchart:~15).
        Re-injected mails are described in section~\ref{tracking
        re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is qmgr
        selecting a mail with a previously unseen queueid for delivery
        (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
        a new data structure will be created.  The mail will be flagged as
        having unknown origins; this flag should be subsequently cleared
        once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        section~\ref{identifying-bounce-notifications} for details.

@@

        This second indicator always occurs for re-injected mail but
        typically it occurs after the first indicator explained above, and
        so rarely introduces re-injection.  This indicator is required to
        tie the parent and child mails together and so is central to the
        process of tracking re-injected mails.

@@

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~23).  Regardless of the state of its children
        the data is cleaned up and entered in the database (flowchart:~24).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~25); for each of those children their data is
        cleaned up and entered in the database, then deleted from the state
        tables.  The parent may also have outstanding children which are
        not yet delivered, in which case (flowchart:~26) the parent must
        wait for those children to be finished with.  As soon as the last
        child is deleted from the state tables the parent will also be
        finished with (flowchart:~27), and deleted from the state tables.

@@

@@

\label{parsing-coverage}

@@

XXX ADD DATES CHECKED FOR ALL PARSERS

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a ``heads up'' for potential trouble spots.\/}
        \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}

    \item [Sawmill Universal Log File Analysis and Reporting] is a
        general purpose commercial product which parses 687 log file
        formats (correct as of 2007/04/15) and produces reports.  I have not
        experimented with it due to its limited data extraction facilities;
        it does have three different sets of data for Postfix (as of
        2007/04/15, one is beta), but they do not appear to be interlinked,
        nor does it save sufficient data for the purposes of this project.
        The source code is available in an obfuscated form only (presumably
        for a fee), and the product is quite expensive, as it requires a
        license for each report which is to be generated. \newline
        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.\/}  \newline
        \url{http://www.enderunix.org/isoqlog/}

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator: the
        data it will extract from an MTA log file is:
        \texttt{time2, email, email\_r, host, host\_r, method, url, code,
        and bytesd.} \newline
        \url{http://awstats.sourceforge.net/} \newline
        \url{http://awstats.sourceforge.net/awstats.mail.html} \newline
        \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}

    \item [Log analyser --- throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur.  \newline
        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}

@@

        \url{http://anteater.drzoom.ch/}

@@

        \url{http://yaala.org/}

    \item [Logparser/Lire] Lire is another general purpose log parser, but
        is developed under the GNU GPL\@.  Like other general purpose log
        parsers it doesn't extract enough data from Postfix logs to be
        worthwhile.  Development also seems to have stalled, with only a
        minor update since October 2004.  \newline
        \url{http://logreport.org/lire/}

@@

        The parsing Logrep performs is extremely basic, counting the number
        of different processes executed, from address, to addresses, number
        of recipients, bytes transferred and delay.  \newline
        \url{http://www.itefix.no/phpws/index.php}

    \item [Log Mail Analyser]~\cite{log-mail-analyser} attempts to
        correlate Postfix log lines by queueid and produce both a CSV file
        and either a Berkeley DB or MySQL database.  It aims to save
        \textit{date and hour, DNS name and IP address host, mail server IP
        address, sender, receiver and e-mail status (sent, rejected)}, so
        its data extraction is reasonably complete.  The parsing is
        hard-coded into the program; extending it requires familiarity with
        a substantial portion of the code.  Producing a CSV file appears to
        be quite limiting, but the intention is to provide a format
        suitable for use with grep(1), awk(1) and the myriad of other Unix
        utilities already available.  Supporting Berkeley DB is also
        limiting, requiring an understanding of the (partially documented)
        database structure and writing a program to extract the required
        data --- it appears to be the least useful of the three output
        formats, though the authors of LMA did provide some queries:
        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
        Unfortunately when the author tested LMA it didn't work properly
        and after exploring the code for some time the author reluctantly
        gave up.

>>>>

% Useful stuff for math mode.
\usepackage{amstext}
% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
\usepackage{url}
%\DeclareUrlCommand\urlwrap{}
%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
% Add the bibliography into the table of contents.
\usepackage[section,numbib]{tocbibind}
% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}

@@

Section~\ref{database schema} partially describes the database schema used by
the parser, explaining in detail the tables used for storing the resulting
data.  The remainder of the schema is laid out in section~\ref{rule
attributes}.

Section~\ref{rules} discusses the parsing rules in detail, describing the
table used to hold the rules and explaining their structure, using an
example rule and sample data it matches successfully against, followed by a
discussion of rule efficiency concerns.

@@

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of 93 log files taken from a mail server handling mail for over
1000 users.  Coverage is described both in terms of the fraction of lines
parsed and the fraction of mails and connections successfully handled by
the parsing algorithm.

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn.  Although every
        vendor implements a different dialect of SQL, the basics are the
        same everywhere.

@@

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules are solely concerned with
identifying a line and extracting data from it, whereas the algorithm's
task is to follow the journey each mail takes through Postfix, piecing the
data extracted by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.

@@

\label{prior art}

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
quite different from this parser: it attempts to present correlated log
data in a form suitable for a systems administrator to search using
grep(1), awk(1) and the myriad of other Unix utilities already available,
producing a CSV file which can subsequently be imported into a
MySQL or Berkeley DB database.  The decision to support both CSV and
Berkeley DB appears to have been a limiting factor:
Comma-Separated Value is a very simple format
where a record is stored in a single line, with fields separated by a comma
or other punctuation symbol.  Problems with CSV files include the
difficulty in escaping separators, providing multiple values for a field
(e.g.\ multiple recipients), adding new fields, and lack of a bundled
schema describing the fields.

Berkeley DB is not an SQL database, it only supports (key, value) pairs; in
the main table the value used is a CSV line, and the key is an integer
referred to by other tables.  The secondary by-sender, by-recipient,
by-date and by-IP tables use the sender/recipient/date/IP as the key, and a
CSV list of integers referring to the main table is used as the value.  In
effect this re-implements SQL foreign keys, but without the full
functionality offered by even the most basic of SQL databases.  It also
requires custom code to search on some combination of the above, though the
authors of LMA did provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT\@.  Berkeley DB appears to be the least useful of the
three output formats

The data stored is limited to time and date, hostname and IP address of
client, sender and recipient addresses, and the SMTP code.  Handling of
multiple recipients, SMTP codes or remote servers\footnote{A single mail
may be sent to multiple remote servers if it was addressed to recipients in
different domains, or Postfix needs to try multiple servers for one or more
recipients.} is not explained in the paper, but the options appear to be:

@@

\section{Database schema}
\label{database schema}

\subsection{Connections table}

\subsection{Results table}

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix logs, and presents
the resulting data in a normalised, easy to use representation.

@@

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\ref{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if it doesn't have any data
worth saving, i.e.\ no rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client IP address or hostname, etc. (action: rejection;
        flowchart:~4); acceptance of one or more mails (action: clone;
        flowchart:~5); or some interleaving of both.

@@

Re-injection due to forwarding sadly lacks explicit log lines of its
own;\footnote{Previously discussed in section~\ref{complications},
complication 3.} re-injection is somewhat awkward to explain because it
overlaps both the mail acceptance and mail delivery sections, so discussion
is deferred to section~\ref{tracking re-injected mail}.

Internally generated mails lack any explicit origin in Postfix 2.2.x and
must be detected using heuristics (see
section~\ref{identifying-bounce-notifications} for details).  Bounce
notifications are the primary example of internally generated mails, though
there may be other types.

@@

The obvious counterpart to mail entering the system is mail leaving the
system, whether it is by deletion, bouncing, or local or remote delivery.
All four are handled in exactly the same way:

@@

    \item Eventually the mail will be delivered, bounced, or deleted by the
        administrator (action: commit; flowchart:~12).  This is the last
        log line for this particular mail (though it may be indirectly
        referred to if it was re-injected).  If it is neither parent nor
        child of re-injection the data is cleaned up and entered in the
        database (flowchart:~14), then deleted from the state tables
        (flowchart:~15).  Re-injected mails are described in
        section~\ref{tracking re-injected mail}.

@@

    \item The indicator which more commonly introduces re-injection is when 
        \texttt{postfix/qmgr}
        selects a mail with a previously unseen queueid for delivery
        (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
        a new data structure will be created.  The mail will be flagged as
        having unknown origins; this flag should be subsequently cleared
        once the origin has been established.  This may also be an
        indicator that the mail is a bounce notification, see
        section~\ref{identifying-bounce-notifications} for details.

@@

        This second indicator always occurs for re-injected mail but
        typically after the first indicator explained above.  This
        indicator is required to tie the parent and child mails together
        and so is central to the process of tracking re-injected mails.

@@

    \item The last alternative is that the mail is a parent mail (action:
        commit; flowchart:~23).  Regardless of the state of its children
        the data is cleaned up and entered in the database (flowchart:~24).
        The parent may have children which are waiting to be entered in the
        database (flowchart:~25); each of those children's data is cleaned
        up and entered in the database, then deleted from the state tables.
        The parent may also have outstanding children which are not yet
        delivered, in which case (flowchart:~26) the parent must wait for
        those children to be finished with.  As soon as the last child is
        deleted from the state tables the parent will also be finished with
        (flowchart:~27), and deleted from the state tables.

@@

\label{parsing coverage}

@@

@@

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a ``heads up'' for potential trouble spots.\/}
        \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}
        \newline (Last checked 2007/08/13.)

    \item [Sawmill Universal Log File Analysis and Reporting] is a general
        purpose commercial product which parses 687 log file formats
        (correct as of 2007/04/15) and produces reports.  I have not
        experimented with it due to its limited data extraction facilities;
        it does have three different sets of data for Postfix (as of
        2007/04/15, one is beta), but they do not appear to be interlinked,
        nor does it save sufficient data for the purposes of this project.
        The source code is available in an obfuscated form only (presumably
        for a fee), and the product is quite expensive, as it requires a
        license for each report which is to be generated. \newline
        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
        \newline (Last checked 2007/08/13.)

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.\/}  \newline
        \url{http://www.enderunix.org/isoqlog/} \newline (Last checked
        2007/08/13.)

@@

        AWStats will produce simple graphs for many different services, but
        that restricts it to supporting the Lowest Common Denominator: the
        data it will extract from an MTA log file is:
        \texttt{time2, email, email\_r, host, host\_r, method, url, code,
        and bytesd.} \newline
        \url{http://awstats.sourceforge.net/} \newline
        \url{http://awstats.sourceforge.net/awstats.mail.html} \newline
        \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
        \newline (Last checked 2007/08/13.)

    \item [Log analyser --- throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur.  \newline
        \url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline
        (Last checked 2007/08/13.)

@@

        \url{http://anteater.drzoom.ch/} \newline (Last checked
        2007/08/13.)

@@

        \url{http://yaala.org/} \newline (Last checked 2007/08/13.)

    \item [Logparser/Lire] Lire is another general purpose log parser, but
        is developed under the GNU GPL\@.  Like other general purpose log
        parsers it doesn't extract enough data from Postfix logs to be
        worthwhile.  Development also seems to have stalled, with only a
        minor update since October 2004.  \newline
        \url{http://logreport.org/lire.html} \newline (Last checked
        2007/08/13.)

@@

        The parsing Logrep performs is extremely basic, counting the number
        of different processes executed, from address, to addresses, number
        of recipients, bytes transferred and delay.  \newline
        \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
        2007/08/13.)

    \item [Log Mail Analyser]~\cite{log-mail-analyser} has already been
        discussed in section~\ref{prior art}.
